---
title: "Geospatial data visualisation in R: bonus challenge"
output: html_document
date: "2022-12-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Note**: only the knitted document (html file) can correctly show the solution of the third exercise.

Libraries:

```{r message=FALSE, warning=FALSE}
library(crosstalk)
library(leaflet)
library(htmltools)
library(DT)
library(sf)
library(here)
```

# Bonus challenge 1

## Prepare data and maps

### Data

Read and prepare data (same code as provided in `20221206_challenges.R`):

```{r read_data}
# natura2000
natura2000 <- st_read(here::here("data",
                                 "20221206",
                                 "20221206_protected_areas.gpkg"))
# occurrences giant hogweed
occs_hogweed <- readr::read_tsv(here::here("data",
                                 "20221206",
                                 "20221206_gbif_occs_hogweed.txt"),
                                na = "")
# transform to sf spatial data.frame
occs_hogweed <- st_as_sf(occs_hogweed,
                         coords = c("decimalLongitude", "decimalLatitude"),
                         crs = 4326)

# count number of "points" in Natura2000 areas
occs_in_areas <- st_contains(natura2000, occs_hogweed)
# get number of points in each polygon as a vector
natura2000$n_occs <- purrr::map_dbl(occs_in_areas, function(x) length(x))
```

## Exercise 1

Create leaflet map:

```{r bonus_challenge_1}
pal <- colorNumeric(
  palette = "viridis",
  domain = natura2000$n_occs)

leaflet(natura2000) %>%
  addTiles() %>%
  addPolygons(popup = ~htmlEscape(NAAM),
              label = ~htmlEscape(n_occs),
              color = ~pal(n_occs)) %>%
  addLegend("bottomright", pal = pal, values = ~n_occs,
            title = "Number of occurrences",
            opacity = 1
  )
```

## Exercise 2

Combine leaflet with data table via crosstalk. Unfortunately, crosstalk doesn't support polygons at the moment (see https://github.com/rstudio/crosstalk/issues/55). Make a leaflet with markers using the centroids of the areas (use `st_centroids()`):

```{r bonus_challenge_2}
natura2000_centroids <- st_centroid(natura2000)
shared_natura2000 <- SharedData$new(natura2000_centroids)

leaflet_map <- leaflet(shared_natura2000) %>%
  addTiles() %>%
  addMarkers(popup = ~htmlEscape(NAAM),
             label = ~htmlEscape(n_occs))
data_table <- datatable(
  shared_natura2000, extensions="Scroller", style="bootstrap", class="compact", width="100%",
  options=list(deferRender=TRUE, scrollY=300, scroller=TRUE)
)
bscols(leaflet_map, data_table)
```

## Exercise 3

Add filter slider:

```{r bonus_challenge_3}
filter_slider(id = "n_occs",
              label = "Number of observations",
              sharedData = shared_natura2000,
              column=~n_occs,
              step=10,
              width = max(natura2000$n_occs))
bscols(leaflet_map, data_table)
```

That's all. I hope you enjoyed it! 
