---
title: "Geospatial data visualisation in R: bonus challenge 2 - leaflet + crosstalk"
output: html_document
date: "2023-08-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Note**: only the knitted document (html file) can correctly show the solution of the third exercise.

Libraries:

```{r message=FALSE, warning=FALSE}
library(crosstalk)
library(leaflet)
library(htmltools)
library(DT)
library(sf)
```

# Bonus challenge 1

## Prepare data

Read and prepare data (same code as provided in `20230831_challenges.R`):

```{r read_data}
# n_moths
n_moths <- readr::read_csv("./data/20230831/20230831_n_moths.txt")

# UTM1km cells of Pajottenland
utm1_pajottenland <- sf::st_read(
  "./data/20230831/20230831_utm1km_Pajottenland.gpkg"
)

# join data
n_moths <- utm1_pajottenland %>%
  dplyr::left_join(n_moths, by = "TAG")
```

### From polygons to points

Generate centroids of the UTM 1x1km cells:

```{r centroids}
n_moths_centroids <- sf::st_centroid(n_moths)
```

### Color palette

Crete the color palette from column `n` with number of moth species:

```{r palette}
pal <- leaflet::colorNumeric(
  palette = "viridis",
  domain = n_moths$n)
```


## Exercise 1

1. Combine leaflet with data table via
[crosstalk](https://rstudio.github.io/crosstalk/index.html). Unfortunately,
crosstalk doesn't support polygons at the moment (see
https://github.com/rstudio/crosstalk/issues/55). So, use the leaflet map with
circle markers you created in the last exercise of BONUS CHALLENGE 1. Hint:
everything you are doing is html. So, use the R package
[DT](https://rstudio.github.io/DT/) to display the data.frame as HTML
"tables".

```{r bonus_challenge_2_1}
shared_n_moths_centroids <- crosstalk::SharedData$new(n_moths_centroids)

# create same leaflet as before but starting from new object wich allows us to
# link data to circles
map_shared_n_moths <- leaflet(shared_n_moths_centroids) %>%
  addTiles() %>%
  addCircleMarkers(popup = ~htmlEscape(TAG),
                   label = ~htmlEscape(n),
                   radius = ~n/10,
                   color = ~pal(n),
                   stroke = FALSE,
                   fillOpacity = 0.5)

# create table with the data
data_table <- datatable(
  shared_n_moths_centroids,
  extensions="Scroller",
  style="bootstrap",
  class="compact",
  width="100%",
  options=list(deferRender=TRUE, scrollY=300, scroller=TRUE)
)

# combine map with table
bscols(map_shared_n_moths, data_table)
```

## Exercise 2

Add filter slider:

```{r bonus_challenge_3}
filter_slider(id = "n",
              label = "Number of moth species",
              sharedData = shared_n_moths_centroids,
              column = ~n,
              step = 5,
              # start from zero to have 0 - 10 - 20 ...ticks
              min = 0,
              width = max(n_moths$n)
)
bscols(map_shared_n_moths, data_table)
```

That's all. I hope you enjoyed it! 
