---
title: "Data visualization"
author: 
- "Damiano Oldoni"
- "you"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: hide
---

# Setup 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = 'styler' # see intermezzo 1
)
```

Load packages:

```{r load-pkgs, message=FALSE, warning=FALSE}
library(tidyverse)    # to do datascience
library(geepack)      # to do modelling
library(INBOtheme)    # to apply INBO style to graphs
library(sf)           # to work with geospatial vector data
library(leaflet)      # to make dynamic maps
library(htmltools)    # to make nice html labels for dynamic maps
library(here)         # to work with paths
```

Notice we have to "reload" tidyverse: the knitting process of a RMarkdown is an universe (_environment_) apart!

Same holds true for `catch_fl`, the cleaned data as output of challenge 1. Again, notice that it is present in your _local_ environment, but won't be  during the knitting process.

Read `catch_fl` containing the cleaned geese catch data:

```{r read-cleaned-data}
path_catch_file <- here("data", "20221213", "20221213_geese_counts_cleaned.txt")
catch_fl <- read_csv(path_catch_file, na = "") 
```

# Introduction

In this section we will show how number of catches varies by year, province and species. 
Both static plots and dynamic maps are generated.

# Static plots

## Show catches  {.tabset .tabset-fade .tabset-pills}

### per province

```{r echo=FALSE}
catch_per_province <- catch_fl %>%
  group_by(province) %>%
  summarize(catched_total = sum(catched, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(catched_total))
ggplot(catch_per_province,
       aes(x = province, y = catched_total)) +
  geom_bar(stat = 'identity') +
  scale_x_discrete(breaks = 2009:2018)
```

### per year

```{r catch-per-year}
catch_per_year <- catch_fl %>%
  group_by(year) %>%
  summarize(catched_total = sum(catched, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(catched_total))
ggplot(catch_per_year,
       aes(x = year, y = catched_total)) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(breaks = 2009:2018)
```

### per year and province

```{r catch-year-pr}
catch_per_year_province <-
  catch_fl %>%
  group_by(year, province) %>%
  summarize(catched = sum(catched, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(catched))
ggplot(catch_per_year_province,
       aes(x = year, y = catched, fill = province)) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(breaks = 2009:2018)
```

## Catch analysis at species level

### Species selection

Before we proceed to analyse the catches at species level, specify the species we are interested to by `commonName`:

```{r species, class.source = "fold-show"}
species <- c(
  "Brandgans",
  "Canadese gans",
  "Grauwe gans",
  "Soepgans",
  "Nijlgans"
)
```

### Catches per per year and species

```{r catch-species}
catch_species <-
  catch_fl %>%
  filter(commonName %in% species) %>%
  group_by(year, commonName) %>%
  summarize(catched_total = sum(catched, na.rm = TRUE)) %>%
  arrange(commonName)
ggplot(catch_species,
       aes(x = year, y = catched_total, fill = commonName)) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(breaks = 2009:2018)
```

### Data modelling

We apply a GEE (generalized estimating equations) model to data from 2010.

```{r apply-gee-model}
model_per_species <-
  map(
    species,
    function(s) {
      dfs <- catch_fl %>%
        filter(commonName == s & year >= 2010) %>%
        arrange(location, year) %>%
        mutate(year = as_factor(as.character(year)),
               location = as_factor(location))
      geeglm(counts ~ 0 + year,
             family = poisson,
             data = dfs,
             waves = year,
             id = location)
    })
names(model_per_species) <- species
overview_model <- map(model_per_species, ~summary(.))
overview_gee <- map2_dfr(
  overview_model,
  names(overview_model), function(model, name) {
    coefficients(model)[,1:2] %>%
      rownames_to_column(var = "year") %>%
      as_tibble() %>%
      mutate(species = name,
             year = str_sub(year, start = 5)) %>%
      # added dplyr:: as select() exists in sf package as well.
      # Some people could get an error at this line otherwise.
      dplyr::select(species, everything()) 
  })
overview_gee <-
  overview_gee %>%
  mutate(
    lwr = exp(Estimate - Std.err),
    upr = exp(Estimate + Std.err),
    Estimate = exp(Estimate)
)
ggplot(overview_gee,
       aes(x = year, y = Estimate, ymin = lwr, ymax = upr)) +
  geom_errorbar(colour = "cyan3") + geom_point(colour = "cyan4") +
  facet_grid(.~species) +
  xlab("year") + ylab("Estimated number of geese per location") +
  theme_inbo(14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 9)) +
  labs(title = "All provinces")

```

# Dynamic maps

We make dynamic _leaflet_ maps of total number of catches per province.

We read the polygons without showing the code (chunk option: `include=FALSE`).

```{r read_plygons, include=FALSE}
pr_fl <- st_read("./data/20221213/20221213_flemish_provinces.gpkg")
```


```{r leaflet-maps}
pr_fl <- pr_fl %>%
  dplyr::left_join(catch_per_province,
                   by = c("TX_PROV_DESCR_NL" = "province"))
bins <- seq(0, 14000, by = 2000)
pal <- colorBin("YlOrRd", domain = pr_fl$catched_total, bins = bins)

labels <- sprintf(
  "<strong>%s</strong><br/>%g catches",
  pr_fl$TX_PROV_DESCR_NL, pr_fl$catched_total
) %>% lapply(HTML)

map_catch_pr <- leaflet(pr_fl) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(catched_total),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"))
map_catch_pr
```
