# Introduction

In this section we will show how number of catches varies by year, province and species. 
Both static plots and dynamic maps are generated.

**note**: by using bookdown we don't have to read the cleaned data from a file as done in challenge 2! The two R Markdown documents are bound together within the same _environment_, so the dataframe `catch_fl` is available in this chapter as well.

# Static plots

## Show catches

Notice that the option `{.tabset}` is not a supported feature of bookdown.

### Catches per province

```{r echo=FALSE, fig.cap = "Total number of catched geese per province from 2009 to 2018"}
catch_per_province <- catch_fl %>%
  group_by(province) %>%
  summarize(catched_total = sum(catched, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(catched_total))
ggplot(catch_per_province,
       aes(x = province, y = catched_total)) +
  geom_bar(stat = 'identity') +
  scale_x_discrete(breaks = 2009:2018)
```

### Catches per year

```{r catch-per-year, fig.cap = "Annual number of catched geese"}
catch_per_year <- catch_fl %>%
  group_by(year) %>%
  summarize(catched_total = sum(catched, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(catched_total))
ggplot(catch_per_year,
       aes(x = year, y = catched_total)) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(breaks = 2009:2018)
```

### Catches per year and province

```{r catch-year-pr, fig.cap = "Catched geese per year from 2009 to 2018. The color indicates the province where catch occurred."}
catch_per_year_province <-
  catch_fl %>%
  group_by(year, province) %>%
  summarize(catched = sum(catched, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(catched))
ggplot(catch_per_year_province,
       aes(x = year, y = catched, fill = province)) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(breaks = 2009:2018)
```

## Catch analysis at species level

### Species selection

Before we proceed to analyse the catches at species level, specify the species we are interested to by `commonName`:

```{r species-to-model, class.source = "fold-show"}
species <- c(
  "Brandgans",
  "Canadese gans",
  "Grauwe gans",
  "Soepgans",
  "Nijlgans"
)
```

### Catches per per year and species

```{r catch-species}
catch_species <-
  catch_fl %>%
  filter(commonName %in% species) %>%
  group_by(year, commonName) %>%
  summarize(catched_total = sum(catched, na.rm = TRUE)) %>%
  arrange(commonName)
ggplot(catch_species,
       aes(x = year, y = catched_total, fill = commonName)) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(breaks = 2009:2018)
```

### Data modelling

We apply a GEE (generalized estimating equations) model to data from 2010.

```{r apply-gee-model}
model_per_species <-
  map(
    species,
    function(s) {
      dfs <- catch_fl %>%
        filter(commonName == s & year >= 2010) %>%
        arrange(location, year) %>%
        mutate(year = as_factor(as.character(year)),
               location = as_factor(location))
      geeglm(counts ~ 0 + year,
             family = poisson,
             data = dfs,
             waves = year,
             id = location)
    })
names(model_per_species) <- species
overview_model <- map(model_per_species, ~summary(.))
overview_gee <- map2_dfr(
  overview_model,
  names(overview_model), function(model, name) {
    coefficients(model)[,1:2] %>%
      rownames_to_column(var = "year") %>%
      as_tibble() %>%
      mutate(species = name,
             year = str_sub(year, start = 5)) %>%
      select(species, everything())
  })
overview_gee <-
  overview_gee %>%
  mutate(
    lwr = exp(Estimate - Std.err),
    upr = exp(Estimate + Std.err),
    Estimate = exp(Estimate)
)
```

```{r show_model_plots, fig.cap = "Modelized catch data for five species from 2010 to 2018."}
ggplot(overview_gee,
       aes(x = year, y = Estimate, ymin = lwr, ymax = upr)) +
  geom_errorbar(colour = "cyan3") + geom_point(colour = "cyan4") +
  facet_grid(.~species) +
  xlab("year") + ylab("Estimated number of geese per location") +
  theme_inbo(14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 9)) +
  labs(title = "All provinces")

```

# Dynamic maps

We first read the geopackage file `20221213_flemish_provinces.gpkg` without including the chunk code (chunk option `include=FALSE`):

```{r include=FALSE}
pr_fl <- st_read("./data/20221213_flemish_provinces.gpkg")
```

We make dynamic _leaflet_ maps of total number of catches per province. **Note**: same data used as in \@ref(catches-per-province).

```{r leaflet-maps, fig.cap="Map showing the total number of catches per province."}

pr_fl <- pr_fl %>%
  dplyr::left_join(catch_per_province,
                   by = c("TX_PROV_DESCR_NL" = "province"))
bins <- seq(0, 14000, by = 2000)
pal <- colorBin("YlOrRd", domain = pr_fl$catched_total, bins = bins)

labels <- sprintf(
  "<strong>%s</strong><br/>%g catches",
  pr_fl$TX_PROV_DESCR_NL, pr_fl$catched_total
) %>% lapply(HTML)

map_catch_pr <- leaflet(pr_fl) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(catched_total),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"))
map_catch_pr
```
