---
title: "Read and visualize ABV occurrence data"
format: 
  html:
    df-print: paged
    toc: true
    toc-location: left
    number-sections: true
    other-links:
      - text: Algemene Broedvogel Monitoring
        href: https://inbo.github.io/abv-rapport/2023/index.html
      - text: GBIF species occurrence cube
        href: https://doi.org/10.15468/dl.b38nw5
editor: source
execute:
  echo: true
  warning: false
  eval: true
date: "today"
---

## Setup

Load libraries:

```{r}
#| message: false
library(tidyverse)    # to do datascience
library(here)         # to build file paths in a project
library(INBOtheme)    # to apply INBO style to graphs
library(sf)           # to work with geospatial vector data
library(plotly)       # to make dynamic plots
library(mapview)      # to make dynamic maps
```


## Introduction

In this document we will:

1. read occurrence cube data
2. explore data
3. visualize data


## Read data

Read **ABV** data from the occurrence cube file `20260331_abv_cube.csv`:

```{r}
#| message: false
#| label: read-abv-data # This adds a label to the code chunk, helpful while rendering and eventually debugging if something goes wrong. R Markdown syntax still works: ```{r read-abv-data}
abv_cube <- read_csv(
  # Specifically for read_* functions of readr package 
  # you can omit the message with column information by
  # using argument `show_col_types = FALSE`
  file = here::here("data", "20260331", "20260331_abv_cube.csv")
)
```

Read the Flemish grid from the geopackage file `20260331_utm_grid.gpkg`:

```{r}
vl_grid <- st_read(
  # Notice that the chunk option `#| message: false` will not work with
  # `st_read()`: the message is still visible. We can use the chunk option 
  # `#|output: false` or we can use the argument `quiet = TRUE` 
  # in `st_read()`.
  dsn = here("data", "20251028", "20251028_utm_grid.gpkg"),
  quiet = TRUE
)
```

## Explore data

This dataset contains data from `{r} min(abv_cube$year)` to `{r} max(abv_cube$year)`
related to `{r} length(unique(abv_cube$specieskey))` species from the
[`{r} unique(abv_cube$family)`](`{r} file.path("https://www.gbif.org/species", unique(abv_cube$familyKey))` ) 
family and their distribution in Flanders based on a grid of 1 km x 1 km.

Preview of the first 30 rows of the dataset:

```{r}
head(abv_cube, n = 30)
```

### Taxonomic information

Species present in the dataset:

```{r}
abv_cube %>% distinct(specieskey, species)
```

### Temporal information

The data are temporally defined at year level. Years present:

```{r}
abv_cube %>% dplyr::distinct(year) %>% arrange(year)
```

### Geographical information

The geographical information is represented by the `mgrscode` column,
which contains the identifiers of the grid cells containing at least one
occurrence of the species.

The dataset contains `{r} length(unique(abv_cube$mgrscode))` unique grid cells.


## Preprocess data

Add geometrical information to the occurrence cube via `mgrscode`, which 
contains the identifiers of the grid cells containing at least one occurrence 
of the species.

```{r}
cells_in_cube <- vl_grid %>%
  dplyr::filter(mgrscode %in% unique(abv_cube$mgrscode)) %>%
  dplyr::select(-c(TAG, Shape_Leng, Shape_Area))
sf_abv_cube <- cells_in_cube %>%
  dplyr::left_join(abv_cube, by = "mgrscode")
```

Final (spatial) dataset:

```{r}
sf_abv_cube %>% head(n = 30)
```


## Visualize data

In this section we will show how the number of occurrences and the number of 
occupied grid cells vary by year and species. Both static plots and dynamic maps are generated.

### Static plots

::: {.panel-tabset}
#### per species

```{r}
#| code-fold: true
#| code-summary: "Show the code"
n_per_species <- sf_abv_cube %>%
  dplyr::group_by(species) %>%
  dplyr::summarize(occurrences = sum(n),
                   grid_cells = n_distinct(mgrscode),
                   .groups = "drop") %>%
  tidyr::pivot_longer(cols = c(occurrences, grid_cells),
                      names_to = "variable",
                      values_to = "n")

ggplot(n_per_species, aes(x = species, y = n)) +
  geom_bar(stat = 'identity') +
  facet_grid(.~variable, scales = "free_y") +
  ggplot2::theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

#### per year

```{r}
#| code-fold: true
#| code-summary: "Show the code"
n_per_year <- sf_abv_cube %>%
  dplyr::group_by(year) %>%
  dplyr::summarize(occurrences = sum(n),
                   grid_cells = n_distinct(mgrscode),
                   .groups = "drop") %>%
  tidyr::pivot_longer(cols = c(occurrences, grid_cells),
                      names_to = "variable",
                      values_to = "n")

ggplot(n_per_year,aes(x = year, y = n)) +
  geom_bar(stat = 'identity') +
  facet_grid(.~variable, scales = "free_y") +
  ggplot2::theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

#### per year and species

```{r}
#| code-fold: true
#| code-summary: "Show the code"
n_occs_per_year_species <-
  sf_abv_cube %>%
  dplyr::group_by(year, species) %>%
  dplyr::summarize(occurrences = sum(n),
                   grid_cells = n_distinct(mgrscode),
                   .groups = "drop") %>%
  tidyr::pivot_longer(cols = c(occurrences, grid_cells),
                      names_to = "variable",
                      values_to = "n")

ggplot(n_occs_per_year_species,
       aes(x = year, y = n, fill = species)) +
  geom_bar(stat = 'identity') +
  facet_grid(.~variable) +
  ggplot2::theme(axis.text.x = element_text(angle = 60, hjust = 1))
```
:::


### Dynamic plots

#### Leaflet dynamic map

We show a map with the distribution of buntings in Flanders. We show the 
total number of occurrences per grid cell. The color of the grid cells is 
based on the number of occurrences. The legend shows the color scale and the 
number of occurrences per grid cell.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
n_occs_per_cell <- sf_abv_cube %>%
  dplyr::group_by(mgrscode) %>%
  dplyr::summarize(
    occurrences = sum(n),
    min_coordinateuncertaintyinmeters = min(mincoordinateuncertaintyinmeters),
    .groups = "drop"
  )

map_abv <- mapview(
  n_occs_per_cell,
  zcol = "occurrences",
  legend = TRUE
)

map_abv
```


#### Plotly yearly abundance

We show a graph with the yearly abundances per species.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
n_occs_per_year <- n_occs_per_year_species |>
  dplyr::filter(variable == "occurrences") |>
  st_drop_geometry()

fig <- plot_ly(
  n_occs_per_year,
  x = ~year,
  y = ~n,
  split = ~species,
  stroke = ~species,
  type = "scatter",
  mode = "lines+markers"
)

fig
```
